Elucidating the Transcriptome of Turkey Hemorrhagic Enteritis Virus

### 2023-12-15 manuscript_thev_transcriptome.Rmd
Corrected the scripts responsible for overestimating the read counts in tables 2a-c.
changed unicode character in ref to "'" to render pdf.

## RESULTS   
**Overview of sequencing data and analysis pipeline outputs**     
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (__Table 1__, __Figure 2a__). From the mapped reads, we identified an overall total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our qPCR assay quantifying the total number of viral genome copies over time (__Figure 2b__). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the final consolidated transcriptome of THEV, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (__Figure 3__). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (http://www.ncbi.nlm.nih.gov/geo) under **accession no. XXXXXX**.

**Changes in THEV splicing profile over time**   
AdV gene expression occurs under exquisite temporal control, supervised by designated promoters for each transcription unit (TU). Each promoter typically produce one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts characteristic of AdVs. To evaluate the activity of each promoter over time, Firstly, `Ballgown` - a program for statistical analysis of assembled transcriptomes, was used to estimate and normalize expression levels of all transcripts in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units over time. Very few unique splice junctions, reads, and transcripts were counted at 4 h.p.i; hence, this time point was excluded in this analysis. This analysis revealed DBP - from the E2 region - to be the most significantly expressed at 12 h.p.i, whereas MLP region transcripts were most abundantly expressed at 24 and 72 h.p.i as expected (__Figure 4a__). Also, our analysis of the FPKM values of transcripts by region showed a similar pattern: the E2 region as the most significantly overexpressed at 12 h.p.i, after which the MLP region assumes predominance (__Figure 4b__).  
Secondly, we estimated relative abundances of all splice junctions for each time point, counting as significantly expressed only those with coverage of at least 1% of the total splice junction reads counted at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were predominantly early region (E1, E2, and E3) junctions, albeit a significant proportion are MLP-derived (__Table 2a__). Junctions from the E2 region were the most abundant at this time point, constituting `r round(most_sig_12, 1)`% of all junctions. The two most abundant junctions at 12 h.p.i were maintained at 24 h.p.i also as the most significantly expressed. However, the MLP-derived junctions were unsurprisingly the most preponderant overall at this time point, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (__Table 2b__). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions were not only the most abundant overall - accounting for `r round(most_sig_72, 1)`% of all junctions reads - but also contained the most significantly expressed individual junctions (__Table 2c__. Also see __Supplementary Tables 1a-c__; __Figure 4c__). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was substantially similar to the pattern seen with all the junctions included (__Figure 4d__).  
Furthermore, we analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides.

**Early Region 1 (E1) transcripts**. This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (__Figure 5__). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the Hyd protein as the 3'-most coding sequence (CDS) if the first start codon (SSC) is skipped. The 5'-most CDS of TRXPT_1 is multi-exonic, producing a 17.9 kilodalton (kDa) protein of 160 residues [amino acids (aa)]. The CDS begins in the first exon, starting at position 211, spans the second exon, and terminates in the third exon at position 2312. From the 5'-most SSC, TRXPT_2 encodes the largest protein in this region ——  a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964 (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. The N-terminus of TRXPT_2 CDS therefore, shares substantial protein sequence similarity with ORF1 but Hyd and TRXPT_2 are not in-frame; hence no protein sequence similarity. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. From our analysis, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2; however, given that studies in MAdVs show that E1 mRNAs share a common 5' and 3' positions, only differing from each other regarding the introns [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are truncated, and the TSS just like the transcription termination site (TTS) are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential remains unaffected. Its 5'-most CDS, beginning at 1965 and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein. This CDS (ORF4) was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003], but our data suggests it is a bona fide ORF. The coding potential of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its coding potential is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. All splice junctions of the transcripts in this region (except the junction for TRXPT_4) have been validated by cloning and Sanger sequencing of cDNA (__Figure 5b__; __supplementary PCR methods__). Finally, during our validation of TRXPT_2, ORF1 was present on the agarose gel and Sanger sequencing results as a bona fide transcript (__supplementary PCR methods__). This is corroborated by 3' Rapid Amplification of cDNA Ends (3'RACE) experiment, which shows transcripts spanning the entire ORF1 and Hyd ORFs without any splicing. The 5'-most CDS of this transcript would encode ORF1. Given that the SSC of the predicted ORF1 is in-frame but downstream of TRXPT_2 SSC, it suggests that the predicted ORF1 CDS is truncated; it shares the same SSC as TRXPT_2, but has a unique STC, albeit it has the same TTS.

@@

### 3' Rapid Amplification of cDNA Ends (3'RACE)  
We performed a rapid amplification of sequences from the 3′ ends of mRNAs (3'-RACE) experiment with total RNA extracted from independently infected MDTC-RP19 cells following the protocol described by Green _et al_ [@Green2019] with modifications. Briefly,

### Computational Analysis of RNA Sequencing Data: Mapping and Transcript characterization
Analysis of our sequence reads were analyzed following a well established protocol described by Pertea _et al_ [@Pertea2016], using `SNAKEMAKE 7.24.0` to drive the pipeline. Briefly, sequencing reads were trimmed with the `FastQC - version 0.11.9` [@FastQC] program to achieve an overall `Mean Sequence Quality (Phred Score)` of 36. Trimmed reads were mapped to the complete sequence of avirulent turkey hemorrhagic enteritis virus strain Virginia (https://www.ncbi.nlm.nih.gov/nuccore/AY849321.1/) and *Meleagris gallopavo* (https://www.ncbi.nlm.nih.gov/genome/?term=Meleagris+gallopavo) using `Hisat2 - version 2.2.1` [@Pertea2016] with default settings without relying on known splice sites. The generated `BAM` files from each infection time-point were filtered for reads mapping to the THEV genome and fed into `StringTie - version 2.2.1` [@Pertea2016] using a `gff3` file from NCBI containing the predicted ORFs of THEV as a guide. A custom script was used to consolidate all transcripts from all time-points without redundancy, generating the final transcriptome of THEV.
\newpage

@@

![](results/r/figures/fig_2.png)
**Figure 2: Increasing levels of THEV over time. a) _Per base coverage of sequence reads mapping to THEV genome by time point_**. The pileup of mRNA reads mapping to THEV genome at the base-pair level for each indicated time point. **b) _One-step growth of THEV (VAS vaccine strain) in MDTC-RP19 cell line_**. After infecting cells at an MOI of 100 GCN/cell, triplicates of harvested infected cells were quantified with an in-house qPCR assay measuring the total copies of THEV genome. There is no discernible increase in virus titer up 12 h.p.i, after which there is a steady increase in virus titer is measured. The virus titer expands exponentially beginning from 48 h.p.i, increasing by orders of magnitude before reaching a plateau at 120 h.p.i. GCN: genome copy number.

>>>>

## RESULTS   
**Overview of sequencing data and analysis pipeline outputs**     
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (__Table 1__, __Figure 2a__). From the mapped reads, we identified an overall total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (__Figure 2b__). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel, ~~and using 3' Rapid Amplification of cDNA Ends (3'RACE) and other methods, we further identified #### unique splice variants~~. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (__Figure 3__). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (http://www.ncbi.nlm.nih.gov/geo) under **accession no. XXXXXX**.

**Changes in THEV splicing profile over time**   
AdV gene expression occurs under exquisite temporal control, supervised by designated promoters for each transcription unit (TU) or region. Each promoter typically produce one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts characteristic of AdVs [@Donovan2020; @Zhao2014]. To evaluate the activity of each promoter over time, Firstly, `Ballgown` - a program for statistical analysis of assembled transcriptomes [@Ballgown] was used to estimate and normalize expression levels of all transcripts in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units over time. Very few unique splice junctions, reads, and transcripts were counted at 4 h.p.i; hence, this time point was excluded in this analysis. This analysis revealed DBP - from the E2 region - to be the most significantly expressed at 12 h.p.i, whereas MLP region transcripts were most abundantly expressed at 24 and 72 h.p.i as expected (__Figure 4a__). Also, our analysis of the FPKM values of transcripts by region showed a similar pattern: the E2 region as the most significantly expressed at 12 h.p.i, after which the MLP region assumes predominance (__Figure 4b__).  
Secondly, we estimated relative abundances of all splice junctions for each time point, counting as significantly expressed only those with coverage of at least 1% of the total splice junction reads counted at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were predominantly early region (E1, E2, and E3) junctions, albeit a significant proportion are MLP-derived (__Table 2a__). Junctions from the E2 region were the most abundant at this time point, constituting `r round(most_sig_12, 1)`% of all junctions. The levels of the two most abundant junctions at 12 h.p.i were maintained also 24 h.p.i as the most significantly expressed. However, the MLP-derived junctions were unsurprisingly the most preponderant overall at this time point, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (__Table 2b__). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions were not only the most abundant overall - accounting for `r round(most_sig_72, 1)`% of all junctions reads - but also contained the most significantly expressed individual junctions (__Table 2c__. Also see __Supplementary Tables 1a-c__; __Figure 4c__). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (__Figure 4d__).  
Furthermore, we analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides.

**Early Region 1 (E1) transcripts**. This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (__Figure 5__). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the Hyd protein as the 3'-most coding sequence (CDS) if the first start codon (SSC) is skipped. The 5'-most CDS of TRXPT_1 is multi-exonic, producing a 17.9 kilodalton (kDa) protein of 160 residues [amino acids (aa)]. The CDS begins in the first exon, starting at position 211, spans the second exon, and terminates in the third exon at position 2312. From the 5'-most SSC, TRXPT_2 encodes the largest protein in this region ——  a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964 (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. The N-terminus of TRXPT_2 CDS therefore, shares substantial protein sequence similarity with ORF1 but Hyd and TRXPT_2 are not in-frame; hence no protein sequence similarity. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. From our analysis, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2; however, given that studies in MAdVs show that E1 mRNAs share a common 5' and 3' positions, only differing from each other regarding the introns [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are truncated, and the TSS just like the transcription termination site (TTS) are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential remains unaffected. Its 5'-most CDS, beginning at 1965 and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein. This CDS (ORF4) was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003], but our data suggests it is a bona fide ORF. The coding potential of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its coding potential is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. All splice junctions of the transcripts in this region (except the junction for TRXPT_4) have been validated by cloning and Sanger sequencing of cDNA (__Figure 5b__; __supplementary PCR methods__). Finally, during our validation of TRXPT_2, ORF1 was present on the agarose gel and Sanger sequencing results as a bona fide transcript (__supplementary PCR methods__). This is corroborated by 3'RACE experiment, which shows transcripts spanning the entire ORF1 and Hyd ORFs without any splicing. The 5'-most CDS of this transcript would encode ORF1. Given that the SSC of the predicted ORF1 is in-frame but downstream of TRXPT_2 SSC, it suggests that the predicted ORF1 CDS is truncated; it shares the same SSC as TRXPT_2, but has a unique STC, albeit it has the same TTS.

@@

### 3' Rapid Amplification of cDNA Ends (3'RACE)  
We performed a rapid amplification of sequences from the 3' ends of mRNAs (3'-RACE) experiment with total RNA extracted from independently infected MDTC-RP19 cells following the protocol described by Green _et al_ [@Green2019] with modifications. Briefly,

### Computational Analysis of RNA Sequencing Data: Mapping and Transcript characterization
Analysis of our sequence reads were analyzed following a well established protocol described by Pertea _et al_ [@Pertea2016], using `SNAKEMAKE 7.24.0` to drive the pipeline. Briefly, sequencing reads were trimmed with the `FastQC - version 0.11.9` [@FastQC] program to achieve an overall `Mean Sequence Quality (Phred Score)` of 36. Trimmed reads were mapped to the complete sequence of avirulent turkey hemorrhagic enteritis virus strain Virginia (https://www.ncbi.nlm.nih.gov/nuccore/AY849321.1/) and *Meleagris gallopavo* (https://www.ncbi.nlm.nih.gov/genome/?term=Meleagris+gallopavo) using `Hisat2 - version 2.2.1` [@Pertea2016] with default settings without relying on known splice sites. The generated `BAM` files from each infection time-point were filtered for reads mapping to the THEV genome and fed into `StringTie - version 2.2.1` [@Pertea2016] using a `gff3` file from NCBI containing the predicted ORFs of THEV as a guide. A custom script was used to consolidate all transcripts from all time-points without redundancy, generating the final transcriptome of THEV.

@@

![](results/r/figures/fig_2.png)
**Figure 2: Increasing levels of THEV over time. a) _Per base coverage of sequence reads mapping to THEV genome by time point_**. The pileup of mRNA reads mapping to THEV genome at the base-pair level for each indicated time point. **b) _Growth of THEV (VAS vaccine strain) in MDTC-RP19 cell line_**. Virus titers were a qPCR assay. There is no discernible increase in virus titer up 12 h.p.i, after which there is a steady increase in virus titer is measured. The virus titer expands exponentially beginning from 48 h.p.i, increasing by orders of magnitude before reaching a plateau at 120 h.p.i. GCN: genome copy number.
